---
title: "Introduction to automatic differentiation, including Laplace approximation and nested approximation"
subtitle: "NIMBLE Alaska ASA Workshop"
author: "NIMBLE Development Team"
date: "June 2026"
format:
  html:
    theme: cosmo
    css: ../styles.css
    toc: true
    code-copy: true
    code-block-background: true
    code-fold: show
    code-tools: true
execute:
  freeze: auto
---
<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>

```{r setup, include=FALSE}
library(nimble)
source(file.path("..", "examples", "DeerEcervi", "load_DeerEcervi.R"), 
       chdir = TRUE)
eval = TRUE
```

# Why derivatives?

Many algorithms require derivatives to work or work well.

* Finding function maximum or minimum (aka optimization)
* Hamiltonian Monte Carlo (HMC; a kind of MCMC sampler)
* Other kinds of MCMC samplers, e.g. Langevin.
* Laplace approximation

    * Starts with an optimization.
    * Then requires matrix of second derivatives (Hessian)
    
* Adaptive Gaussian quadrature (builds upon Laplace approx.)
* Nested approximations (INLA, ELGM (Stringer et al.))

# How do computers get derivatives?

Example function: 

* $y = e^{x^2}$. 
* $\frac{dy}{dx} = 2x e^{x^{2}} = 2xy$. 

Derivative methods:

1. Hard-coded analytic function: Developer codes `2 * x * exp(x^2)`
2. Symbolic algebra: Code processes `exp(x^2)` into `2 * x * exp(x^2)`
3. Finite-element approximation: $\frac{dy}{dx} \approx \frac{e^{(x+\delta)^{2}} - e^{{x^{2}}}}{\delta}$, for some small value of $\delta$ (hard to choose in general).  Code calls function once with input $x$ and once with input $x+\delta$.
4. **Automatic (or "Algorithmic") Differentiation:**: Carry derivative information through all math operators, i.e. `^` and `exp` in $e^{{x^{2}}} =$ `exp(x^2)`.

# Automatic Differentiation (AD)

Steps to compute $y = e^{x^2}$ (**value**) for x = 3.0:

1. $x^{2} = 3^2 \rightarrow 9$.
2. $e^{x^{2}} = e^9 \rightarrow 8103.1$.

Steps to compute $\frac{dy}{dx}$ (using chain rule): *Forward mode*, i.e. inner derivatives first.

1. $\frac{d(x^{2})}{dx} = 2x = 2 \times 3 \rightarrow 6$.
2. $\frac{d(e^{x^2})}{dx} = \frac{d(e^{x^2})}{d(x^{2})} \times \frac{d(x^{2})}{dx} = e^{9} \times 6 = 8103.1 \times 6 \rightarrow 48618.6$.

Steps to compute $\frac{dy}{dx}$ (using chain rule): *Reverse mode*, i.e. outer derivatives first.

1. $\frac{d(e^{x^2})}{d(x^{2})} = e^9 \rightarrow 8103.1$.
2. $\frac{d(e^{x^2})}{dx} = \frac{d(e^{x^2})}{d(x^{2})} \times \frac{d(x^{2})}{dx} = 8103.1 \times 2x = 8103.1 \times 6 \rightarrow 48618.6$

# Some general points about AD 

* AD works by stepping through the same sequence of operations multiple times, tracking different information each time.
* AD can sometimes be *more efficient* than using the derivative equations explicitly.  E.g. $e^{9} = 8103.1$ was calculated and can be re-used.
* AD can be iterated to obtain arbitrary-order derivatives.  Lower-order derivatives must be done before higher-order derivatives.
* Forward and reverse modes are each more efficient for different kinds of problems, depending on number of inputs and outputs (details not discussed here).
* Operations such as matrix multiplication, matrix inversion, matrix solving or others can be implemented with matrix AD operations (rather than relying on potentially long sequences of scalar operations, which can be slow for AD).

# NIMBLE adopts the CppAD C++ library for AD 

* Huge credit to [<span style="color:blue">CppAD</span>](https://coin-or.github.io/CppAD/doc/cppad.htm)
* Written by Brad Bell at the University of Washington.
* CppAD is also used by [<span style="color:blue">Template Model Builder (TMB)</span>](https://kaskr.github.io/adcomp/Introduction.html).
* CppAD is general and efficient.  It can provide arbitrary-order derivatives.
* We have written some specialized handling for some cases ("atomic" operators, in CppAD terminology).
* CppAD can trace derivatives through arbitrary function calls.
* NIMBLE's use of CppAD involves deep changes in everything, not just calling a few extra functions.
* NIMBLE with CppAD can give derivatives of arbitrary model calculations and/or nimbleFunction calculations.
* Care is required for any conditional steps in functions, such as if-then-else or for-loops.  Normally, conditional branching and looping is "baked in" to AD calculations.  This is discussed more later.

# Some terminology for AD in NIMBLE (adapted from CppAD)

* Orders 0, 1, and 2 are called "Value", "Jacobian" (or "Gradient"), and "Hessian", respectively.
* We say that operations are *recorded* onto a *tape*, which is replayed to obtain values (0th order) and/or derivatives (1st and 2nd order).
* We can take *derivatives of derivatives*.  This involves making a *tape of the replaying of a tape*.

    * We call this **double-taping** or **meta-taping**.

* Double-taping can improve efficiency.
* Double-taping can give derivatives of order > 2. 

# First example: Derivatives in a nimbleFunction

```{r, eval=eval}
nf1 <- nimbleFunction(
  setup = TRUE,
  methods = list(
    foo = function(x = double()) {
      return(exp(x^2))
      returnType(double())
    },
    derivs_foo = function(x=double()) {
      d <- derivs(foo(x), wrt = "x", order = 0:2)
      return(d)
      returnType(ADNimbleList())
    }
  ),
  buildDerivs = 'foo'
)
```

# First example: Derivatives in a nimbleFunction

Make an object (instance) of nf1:
```{r, eval=eval}
nf1obj <- nf1()
```

Uncompiled execution:
```{r, eval=eval}
nf1obj$derivs_foo(3)
nf1obj$foo(3) # for comparison of value
```

# First example: Derivatives in a nimbleFunction
Compiled execution:
```{r, eval=eval}
cnf1obj <- compileNimble(nf1obj)
cnf1obj$derivs_foo(3)
```

# Understanding derivative outputs:

* Define $n$ = length of inputs and $m$ = length of outputs.
* **value** is a vector of length $m$.
* **jacobian** is an $m \times n$ matrix ($m$ rows and $n$ columns).
* **hessian** is an $n \times n \times m$ array.
* Notice the output index is the *first* index in *jacobian* but the *last* index in *hessian*.
* First two indices of *hessian* are interchangeable due to symmetry of partial derivatives.
* You can omit lower-orders (e.g., use `order = 1` for the jacobian only, omitting the value), but lower orders *will be calculated* in order to obtain higher orders.
* `ADNimbleList()` is a pre-defined *nimbleList* type with elements `value`, `jacobian`, and `hessian`.

# Second example: Derivatives of model log probabilities

Make a simple model.

```{r, eval=eval}
model_code <- nimbleCode({
  y ~ dnorm(mean = mu, sd = sigma)
  mu <- intercept + slope * x
  intercept ~ dnorm(0, sd = 1000)
  sigma ~ dunif(0, 10)
})
m <- nimbleModel(model_code,
                 inits = list(intercept = 1.2, slope = 0.5,
                              sigma = 0.7),
                 data = list(y = 6),
                 constants = list(x = 10),
                 buildDerivs = TRUE, # ALLOW DERIVATIVES FOR THIS MODEL
                 )
```

# Second example: Derivatives of model log probabilities

Write a nimbleFunction to get derivatives of model$calculate.

```{r, eval=eval}
nf2 <- nimbleFunction(
  setup = function(model, nodes) {},
  methods = list(
    derivs_log_prob = function() {
      d <- derivs(model$calculate(nodes),
                  wrt=c("intercept", "slope", "sigma"),
                  order = 0:2)
      return(d)
      returnType(ADNimbleList())
    }
  )
)
```

# Second example: Derivatives of model log probabilities

Use it uncompiled.

```{r, eval=eval}
calc_nodes <- m$getDependencies(c("intercept", "slope", "sigma"))
calc_nodes
nf2obj <- nf2(m, calc_nodes)
nf2obj$derivs_log_prob()
```

# Second example: Derivatives of model log probabilities

Use it compiled.

```{r, eval=eval}
comp <- compileNimble(m, nf2obj)
comp$nf2obj$derivs_log_prob()
```

# More about understanding outputs and testing

* Uncompiled derivatives use finite-element methods from packages `numDeriv` and/or `pracma`.

    * These might not match compiled derivatives exactly due to finite-element approximation.
    * Uncompiled execution can be slow for large calculations.

* The first compiled call to `derivs` (aka `nimDerivs`) records the AD tape, so it is much slower than later calls.
* Always test compiled derivatives on new values after the first call.  

    * This checks that values have not been (incorrectly) "baked in" to the AD tape. 
    * This was not illustrated in first two examples.
    
* Model inputs are ordered to match the `wrt` argument, flattened and concatenated.

    * This is essential to use `jacobian` and `hessian` results correctly.

# Third example: Derivatives of model log probabilities with other calculations.

Use the same model as above.  For good measure, make a new copy.
```{r, eval=eval}
m2 <- m$newModel()
```

Say we will provide a vector of `(intercept, slope, log_sigma)` and we need to calculate `sigma = exp(log_sigma)` before using it in the model.
```{r, eval=eval}
nf3 <- nimbleFunction(
  setup = function(model, param_nodes, calc_nodes) {
    # param_nodes will be c("intercept", "slope", "sigma")
    updateNodes <- character() # more about these later
    constantNodes <- "y"
  },
  run = function(param_values = double(1)) {
    # param_values will be (intercept, slope, log_sigma)
    param_values[3] <- exp(param_values[3])
    values(model, param_nodes) <<- param_values # The ONLY valid way to put values in a model for AD
    log_prob <- model$calculate(calc_nodes)
    return(log_prob)
    returnType(double())
  },
  methods = list(
    derivs_run = function(param_values = double(1)) {
      wrt <- 1:length(param_values) # indices, not names
      ans <- derivs(run(param_values), wrt = wrt,
                    order = 0:2,
                    updateNodes = updateNodes,
                    constantNodes = constantNodes,
                    model = model)
      return(ans)
      returnType(ADNimbleList())
    }
  ),
  buildDerivs = 'run'
) 
```

# Third example: Derivatives of model log probabilities with other calculations.

Make an object:
```{r, eval=eval}
param_nodes <- c("intercept", "slope", "sigma")
calc_nodes <- m2$getDependencies(param_nodes, self = FALSE) # omit priors, just to make it more interesting
calc_nodes
nf3obj <- nf3(m2, param_nodes, calc_nodes)
```

Use it uncompiled: 

```{r, eval=eval}
x <- c(1.2, .5, log(0.7))
nf3obj$derivs_run(x)
```

# Third example: Derivatives of model log probabilities with other calculations.

Use it compiled:

```{r, eval=eval}
comp2 <- compileNimble(m2, nf3obj)
x_tape <- c(1.4, .2, log(0.4))
comp2$nf3obj$derivs_run(x_tape)
x <- c(1.2, .5, log(0.7))
comp2$nf3obj$derivs_run(x)
```

# Understanding the results

* `values(model, nodes) <<- x` is the *only* way to put values into model nodes that will be recorded on the AD tape.
* `model$x <<- value` and `model[[node]] <<- value` will not be recorded and are errors.  They may be supported in the future. 
* In example 3, only derivatives with respect to (wrt) the third input differ from example 2, because example 3 includes `sigma = exp(log_sigma)`.
* `updateNodes` and `constantNodes` are essential for correct use of the model and correct results.  
* Other points not covered yet include:

    * `reset` argument to `derivs` (aka `nimDerivs`), which resets the AD tape.
    * `wrt` argument can be indices or names.
    * What variables and/or steps get "baked in" to AD tapes.
    * Omitting variables from derivative tracking.
    * Double-taping.
    * System for generic parameter transformations.
    
# Summary of ways to use AD 

* For calculations outside of a model: `nimDerivs(foo(x), ...)` , i.e. "nimDerivs of a function".
* For model log probabilities with no other steps: `nimDerivs(model$calculate(...), ...)`, i.e. "nimDerivs of model calculate".
* For model log probabilities amid other steps: `nimDerivs(foo(x),...)` where `foo(x)` contains `model$calculate(...)`, i.e. "nimDerivs of a function containing model calculate".

# Laplace approximation

Laplace approximation:

* data: $y$
* random effect: $x$
* parameters: $\theta$
* Want marginal probability of data: $\int P(y | x, \theta) P(x | \theta) dx$.
* Approximate by:

  * Find $x^{*}$ that maximizes $P(y | x, \theta) P(x | \theta)$ with respect to $x$.
  * Compute the Hessian of $\log(P(y | x, \theta) P(x | \theta))$ at $x = x^{*}$.
  * Do a calculation with $P(y | x^{*}, \theta) P(x^{*} | \theta)$ and the Hessian.
  
# Laplace: Use Deer E. cervi GLMM

We've used this previously. Here is the code in case we need to look at it:

```{r, eval=FALSE}
DEcode <- nimbleCode({
  for(i in 1:2) {
    # Priors for intercepts and length coefficients for sex = 1 (male), 2 (female)
    sex_int[i] ~ dnorm(0, sd = 1000)
    length_coef[i] ~ dnorm(0, sd = 1000)
  }

  # Priors for farm random effects and their standard deviation.
  farm_sd ~ dunif(0, 20)
  for(i in 1:num_farms) {
    farm_effect[i] ~ dnorm(0, sd = farm_sd)
  }

  # logit link and Bernoulli data probabilities
  for(i in 1:num_animals) {
    logit(disease_probability[i]) <-
      sex_int[ sex[i] ] +
      length_coef[ sex[i] ]*length[i] +
      farm_effect[ farm_ids[i] ]
    Ecervi_01[i] ~ dbern(disease_probability[i])
  }
})

DEconstants <- list(num_farms = 24,
                    num_animals = 826,
                    length = DeerEcervi$ctrLength,
                    sex = DeerEcervi$Sex,
                    farm_ids = DeerEcervi$farm_ids)
```
  
# Laplace for GLMM

Make Deer E. cervi model with derivative tools built.

```{r, eval=eval}
DEmodel <- nimbleModel(DEcode,
                       data = DEdata,
                       constants = DEconstants,
                       inits = DEinits_vals,
                       buildDerivs = TRUE)

cDEmodel <- compileNimble(DEmodel)
```

# Laplace for GLMM using nimbleQuad 

```{r, eval=eval}
library(nimbleQuad)
DE_Laplace <- buildLaplace(DEmodel)
cDE_Laplace <- compileNimble(DE_Laplace, project = DEmodel)
print(system.time(laplace_fit <- runLaplace(cDE_Laplace)))
laplace_fit$summary$params

```

# Laplace for GLMM

Since this is a GLMM, we can compare to a specialized implementation of Laplace approximation.  We'll use `lme4`.

```{r, eval=eval}
library(lme4)
lme4_fit <- glmer(Ecervi_01 ~ ctrLength*fSex + (1 | fFarm), family = binomial, data = DeerEcervi)
summary(lme4_fit)
```

# Laplace for GLMM

How do the parameters match between `nimble` and `lme4` results?

`lme4` is using treatment contrasts (Sex1 is the "reference" group and Sex2 parameters are differences from Sex1).

* First parameter from `nimble` is the Sex1 intercept, which is the `(Intercept)` from `lme4`.
* Second parameter from `nimble` is the Sex2 intercept, which is `(Intercept)` + `fSex2` from `lme4`.
* Third parameter from `nimble` is the Sex1 coefficient for `ctrLength`.
* Fourth parameter from `nimble` is the Sex2 coefficient for `ctrLength`, which is `ctrLength` + `ctrLength:fSex2` from `lme4`.
* Fifth parameter from `nimble` is the farm-effect standard deviation.

# INLA-like deterministic nested approximation

One can use Laplace to approximately marginalize over latent nodes in the context of a deterministic Bayesian approximation.

1. (Approximately) marginalize over latent nodes using Laplace
2. Find posterior mode for (hyper)parameters and create a grid for quadrature
3. Find approximate marginal posterior at each grid point and approximate univariate hyperparameter marginals via quadrature approaches
4. Sample from mixture of MVN distributions for the latent nodes, mixing over the hyperparameter values at the grid points (INLA also derives analytic univariate marginals)

# Nested approximation for GLMM

```{r}
library(nimbleQuad)
approx <- buildNestedApprox(DEmodel)
capprox <- compileNimble(approx, project = DEmodel, resetFunctions = TRUE)
approx_fit <- runNestedApprox(capprox)
approx_fit
smp <- approx_fit$sampleLatents(500)
cbind(apply(smp[ ,1:8], 2, mean), apply(smp[,1:8], 2, sd))

laplace_fit$summary$randomEffects[1:4, ]
```

Results are generally pretty similar between Laplace and the nested approximation here (a bit of discrepancy for the random effects sd), although note that Laplace is frequentist and the nested approximation is Bayesian.

With the nested approximation one would generally included fixed effects in the latent nodes but one could also (manually) include them as parameters, which is how Laplace treats them.

